---
title: "Building My Analysis Dataset"
subtitle: "Gender Gaps in Internet Access and Economic Outcomes"
author: "Shukria Alavi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# My research question

Does a smaller gender gap in internet access go together with better economic
outcomes for women in developing countries?

I look at three outcomes: GDP per capita, female labour force participation,
and female secondary school enrollment.

**What I want to know at this stage:**

1. How big is the internet gender gap, and how much does it vary between
   countries?
2. Do I have enough complete data to compare it against my three outcomes?



# Step 1: Load the three datasets

```{r load}
internet  <- read_csv("data_raw/source1_internet_gender_findex.csv",
                      show_col_types = FALSE)
outcomes  <- read_csv("data_raw/source2_outcomes_worldbank.csv",
                      show_col_types = FALSE)
inequality <- read_csv("data_raw/source3_gii_undp.csv",
                       show_col_types = FALSE)
```

## Dataset 1: Internet use by gender

**Source:** World Bank Global Findex 2025. Gallup surveyed adults aged 15 and
over in each country between May and December 2024 and asked whether they use
the internet. Every country was surveyed in the same year with the same
question, which is why I use this as my main measure.

The published file is 8,577 rows by 438 columns covering six survey years and
every breakdown. `00_download_sources.R` narrows it to 2024, the gender
breakdown, and the internet question, and converts the values from proportions
to percentages.

This gives me my independent variable.

```{r look1}
dim(internet)
names(internet)
head(internet, 3)
```

## Dataset 2: Economic and education outcomes

**Source:** World Bank World Development Indicators. This gives me my three
outcome variables plus two control variables. For each country I took the most
recent value between 2021 and 2024, and kept a column recording which year it
came from.

```{r look2}
dim(outcomes)
names(outcomes)
head(outcomes, 3)
```

## Dataset 3: Gender Inequality Index

**Source:** UNDP Human Development Report 2025. The Gender Inequality Index
combines maternal mortality, adolescent births, seats in parliament, education
and labour force participation into one score from 0 to 1, where higher means
more unequal. I use it as a control.

```{r look3}
dim(inequality)
names(inequality)
head(inequality, 3)
```


# Step 2: Checking the datasets before joining

If the column I join on does not match across files, countries disappear
without any error message. So I check first.

*Does any country appear twice?*

```{r check-dup}
sum(duplicated(internet$country_code))
sum(duplicated(outcomes$country_code))
sum(duplicated(inequality$country_code))
```

All zero, so each country appears once in each file.

**Which countries are in one file but not another?**

`anti_join` shows rows on the left with no match on the right.

```{r check-anti}
anti_join(internet, outcomes, by = "country_code") %>%
  select(country_code, country_name)

anti_join(internet, inequality, by = "country_code") %>%
  select(country_code, country_name)
```

Every country in my internet data is also in the World Bank file. Three are
missing from the UNDP file, so they will have no Gender Inequality Index.

**Are country names spelled the same way?**

```{r check-names}
name_check <- internet %>%
  select(country_code, name_worldbank = country_name) %>%
  inner_join(inequality %>% select(country_code, name_undp = country_name),
             by = "country_code")

name_check %>% filter(name_worldbank != name_undp)
```

```{r count-names}
name_check %>%
  summarise(total = n(),
            same_spelling      = sum(name_worldbank == name_undp),
            different_spelling = sum(name_worldbank != name_undp))
```

**This is the most important check on the page.** 14 countries are spelled
differently between the two organisations. The World Bank writes
"Egypt, Arab Rep." where the UNDP writes "Egypt", and "Kyrgyz Republic" where
the UNDP writes "Kyrgyzstan".

If I had joined on country name, those 14 countries would have silently
dropped out and I would never have known. This is why I join on the
three-letter **country code**, which is standardised across organisations.


# Step 3: Joining

`left_join` keeps every row on the left and adds matching columns from the
right. Where there is no match it fills in `NA` instead of dropping the row.

I start from the internet data, because a country with no gender gap measure
is no use to me.

```{r join}
data <- internet %>%
  left_join(outcomes,   by = "country_code") %>%
  left_join(inequality, by = "country_code")

dim(data)
```

The join added **columns**, not rows.

```{r join-check}
nrow(internet)                    # before
nrow(data)                        # after
sum(duplicated(data$country_code))  # should be 0
```

The join brought in three copies of the country name, one per file. I keep one.

```{r tidy-names}
data <- data %>%
  select(-country_name.y, -country_name) %>%
  rename(country = country_name.x)
```


# Step 4: Cleaning

Each decision below says what I did and why.

**Build the gender gap variable**

The gap is not in any dataset. I calculate it from the male and female columns.

```{r make-gap}
data <- data %>%
  mutate(
    gap_diff  = internet_men - internet_women,   # percentage points; + means men ahead
    gap_ratio = internet_women / internet_men    # 1.0 means equal
  )
```

**Droping Kosovo**

Kosovo is missing female labour force participation, overall internet use, and
the Gender Inequality Index. That is most of what I need. R would exclude it
from nearly every model anyway, so removing it on purpose keeps my stated
sample size honest.

```{r drop-kosovo}
data <- data %>% filter(country != "Kosovo")
nrow(data)
```

**Add a logged GDP column**

GDP per capita is skewed, with a few countries much richer than the rest.
Taking the log evens it out, which is standard before using it in a model.

```{r log-gdp}
data <- data %>% mutate(log_gdp = log(gdp_pc_ppp))
```

**Keep the columns I actually use**

```{r select-cols}
data <- data %>%
  select(country_code, country,
         internet_women, internet_men, gap_diff, gap_ratio,
         gdp_pc_ppp, log_gdp, flfp, sec_enroll_fem_gross,
         internet_total, urban_pct, gender_inequality_index)

dim(data)
names(data)
```

**What I did NOT do**

- I did not remove outliers. Pakistan has the widest gap and the data is
  correct, so it stays.
- I did not fill in missing values. I report them instead.


# Step 5: Check the cleaned data

```{r check-final}
nrow(data)
sum(duplicated(data$country_code))
```

```{r check-missing}
data %>%
  summarise(across(c(gap_diff, gdp_pc_ppp, flfp, sec_enroll_fem_gross,
                     internet_total, urban_pct, gender_inequality_index),
                   ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing") %>%
  arrange(desc(missing))
```




# Step 6: Look at each variable on its own

**The internet gender gap**

```{r gap}
summary(data$gap_diff)

ggplot(data, aes(x = gap_diff)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Internet gender gap",
       x = "Male % minus female % (percentage points)",
       y = "Number of countries")
```

```{r gap-extremes}
data %>% arrange(desc(gap_diff)) %>%
  select(country, internet_women, internet_men, gap_diff) %>% head(5)

data %>% arrange(gap_diff) %>%
  select(country, internet_women, internet_men, gap_diff) %>% head(5)
```



**GDP per capita**

```{r gdp}
summary(data$gdp_pc_ppp)

ggplot(data, aes(x = gdp_pc_ppp)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  labs(title = "GDP per capita", x = "PPP dollars", y = "Number of countries")
```

**Female labour force participation**

```{r flfp}
summary(data$flfp)

ggplot(data, aes(x = flfp)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  labs(title = "Female labour force participation",
       x = "% of women aged 15+", y = "Number of countries")
```

**Female secondary enrollment**

```{r enroll}
summary(data$sec_enroll_fem_gross)

ggplot(data, aes(x = sec_enroll_fem_gross)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  labs(title = "Female secondary enrollment",
       x = "% gross", y = "Number of countries")
```


# Step 7: Save

```{r save}
if (!dir.exists("data_clean")) dir.create("data_clean")
write_csv(data, "data_clean/analysis_data.csv")
```

**Where I am**

- I have **89 countries** with a gender gap measure and most outcomes.
- The gap ranges from women being ahead by 8 points to men being ahead by 44.
- 15 countries have no female secondary enrollment figure, which is my weakest
  outcome variable.


