---
title: "Cleaning My Data"
author: "Shukria Alavi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

This week I cleaned the dataset I built earlier so it is ready to analyse.

# Loading the data

```{r load}
data <- read_csv("research/gender_gap_crosssection_96.csv",
                 show_col_types = FALSE)

dim(data)
```

# Looking at what I have

Before changing anything I want to see what is actually in the file.

```{r look}
names(data)
```

```{r missing}
# How many missing values in the columns I care about?
sum(is.na(data$gap_diff))
sum(is.na(data$gdp_pc_ppp))
sum(is.na(data$flfp))
sum(is.na(data$sec_enroll_fem_gross))
sum(is.na(data$gii))
```


```{r missing-who}
# Which countries are missing enrollment?
data %>%
  filter(is.na(sec_enroll_fem_gross)) %>%
  select(country)
```

# Cleaning

## Step 1: Drop Kosovo

Kosovo is missing female labour force participation, overall internet use,
and the Gender Inequality Index. That is most of what I need, so I remove it
rather than letting R quietly drop it later.

```{r drop-kosovo}
data <- data %>%
  filter(country != "Kosovo")

nrow(data)
```

## Step 2: Keep only the columns I need

The file has 37 columns and I only use about a dozen. I keep those and give
the enrollment column a shorter name.

```{r keep-columns}
data <- data %>%
  select(
    country,
    region,
    income,
    source,
    ref_year,
    internet_female,
    internet_male,
    gap_diff,
    gap_ratio,
    gdp_pc_ppp,
    flfp,
    enroll_fem = sec_enroll_fem_gross,
    internet_total,
    urban_pct,
    gii
  )

dim(data)
```

## Step 3: Add a logged GDP column

GDP per capita is skewed, with a few countries far richer than the rest.
Taking the log evens it out, which is standard before using it in a model.

```{r log-gdp}
data <- data %>%
  mutate(log_gdp = log(gdp_pc_ppp))
```

## Step 4: Turn region and income into categories

Right now these are text. Making them factors tells R they are categories.

```{r factors}
data <- data %>%
  mutate(
    region = factor(region),
    income = factor(income,
                    levels = c("Low income",
                               "Lower middle income",
                               "Upper middle income"))
  )
```

# Checking the cleaned data

```{r check}
dim(data)

summary(data$gap_diff)

table(data$income)
```

```{r check-missing}
# Missing values after cleaning
sum(is.na(data$gap_diff))
sum(is.na(data$gdp_pc_ppp))
sum(is.na(data$flfp))
sum(is.na(data$enroll_fem))
sum(is.na(data$gii))
```

# A first look at the data

## How big is the gender gap?

```{r histogram}
ggplot(data, aes(x = gap_diff)) +
  geom_histogram(bins = 20, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "Internet gender gap",
       x = "Male % minus female % (percentage points)",
       y = "Number of countries")
```

```{r extremes}
# The five biggest gaps
data %>%
  arrange(desc(gap_diff)) %>%
  select(country, internet_female, internet_male, gap_diff) %>%
  head(5)
```


## Does the gap relate to female labour force participation?

```{r scatter}
ggplot(data, aes(x = gap_diff, y = flfp)) +
  geom_point(colour = "steelblue") +
  geom_smooth(method = "lm") +
  labs(x = "Internet gender gap (percentage points)",
       y = "Female labour force participation (%)")
```


# Saving the cleaned data

```{r save}
write_csv(data, "research/cleaned_data.csv")
```

